---
title: "Word order in UD"
author: ''
date: "10/15/2024"
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

This is the code used for Section 4.1.1 of the paper 'Why we need a gradient approach to word order'. https://doi.org/10.1515/ling-2021-0098 

# Read the data

The data originates from the corpora of the Universal Dependencies. 
```{r message=FALSE, warning=FALSE}
# read relevant packages
library(tidyverse)
library(zoo)
# remove scientific notations such as 10000=1e04
options(scipen=999,stringsAsFactors = FALSE) 

# have the list of languages
Languages <- c("Arabic","Basque", "Bulgarian", "Catalan" , "Chinese", "Croatian", "Danish", "Dutch", "English", "Estonian", "Finnish", "French", "Galician", "German", "Hebrew", "Hindi", "Indonesian", "Italian", "Japanese", "Korean", "Latin", "Latvian", "Norwegian", "Persian", "Polish","Portuguese","Romanian", "Russian", "Slovenian", "Spanish", "Swedish", "Ukrainian")

# take a smaller list if needed
Languages <- c("Basque","Catalan")

# open an empty table to store the output
data <- NULL %>% as.data.frame()

# extract data for the list of languages
for(z in c(1:length(Languages))){
  # take all the files in the folder of that language
  files <- list.files(paste("data_raw/UD/",Languages[z],"/",sep=""))
  # create and view an object with file names and full paths
  f <- file.path("data_raw/UD/",Languages[z], files)
  d <- lapply(f, FUN = function(files){read.delim(files,
                                                  header = FALSE, 
                                                  comment.char = "#", 
                                                  stringsAsFactors = FALSE)})
  # combine all the files that were read
  merge.data <- plyr::rbind.fill(d)
  # add the language annotations
  merge.data <- merge.data %>%
    mutate(Language = Languages[z])
  # merge with the entire data
  data <- rbind(data, merge.data)
}
# remove not used vectors
rm(merge.data, d)

# arrange the columns of the table
data <- data %>%
  select(ID_word = V1, Tag = V6, POS = V4, Lemma = V3, 
         Dependency = V7, Role = V8, Language) 

# print the data as a table
data %>% write.csv("data_raw/UD.csv",
                   row.names = FALSE,
                   fileEncoding = "UTF-8")

# visual check
glimpse(data)
```


# Subject and Object

We first need to re-arrange the UD data by adding sentence IDs.
```{r message=FALSE, warning=FALSE}
# if your computer is slow, can read the output file from the previous chunk
data <- read.csv("data_raw/UD.csv")

# adding start and end of sentences
data <- data %>%
  # change IDs to numeric
  mutate(ID_word = as.numeric(ID_word)) %>%
  # add gap of IDs between consecutive pair of words
  mutate(diff = ID_word - lag(ID_word, default = first(ID_word))) %>%
  # change NAs to 0s if needed
  replace(is.na(.), 0) %>%
  # add labels 
  mutate(ID_sentence = case_when(diff < 0 ~ "New_sentence",
                                 diff >= 0 ~ "In")) 

# change new sentence markers to sentence number
data$ID_sentence[which(data$ID_sentence == "New_sentence")] <- 2:(length(data$ID_sentence[which(data$ID_sentence == "New_sentence")])+1)
# manually add the start of the first sentence
data$ID_sentence[1] <- 1

# arrange the data
data <- data %>% 
  # change the sentence ID to numeric
  mutate(ID_sentence = as.numeric(ID_sentence)) %>%
  # remove the diff column
  select(-diff)

# change NAs to the sentence ID
data$ID_sentence <- na.locf(data$ID_sentence)
# print the data as a table
data %>% write.csv("data_tidy/UD.csv",
                   row.names = FALSE,
                   fileEncoding = "UTF-8")

# visual check
glimpse(data)
```


for each sentence extract the order of SVO (this chunk for sampling).
```{r}
# if needed, read the output from the previous chunks
data <- read.csv("data_tidy/UD.csv")

# open an empty table
data.sample <- NULL %>% as.data.frame()

for(w in seq(20,2000,20)){
tmp <- data %>%
  filter(Role %in% c("nsubj","obj")) %>%
  filter(POS == "NOUN") %>%
  # take relevant columns
  select(ID_sentence, ID_word, Dependency, Role, Language) %>%
  # change columns to numeric if needed
  mutate(Dependency = as.numeric(Dependency),
         ID_word = as.numeric(ID_word)) %>%
  #can play with this setting to see if want to take samples by roles too
  group_by(Language,Role) %>%
  sample_n(size = w, replace = T) %>%
  ungroup() %>%
  mutate(position = ID_word-Dependency) %>% 
  mutate(position = case_when(position < 0 ~ "before_verb",
                              position > 0 ~ "after_verb"))  %>%
  # add the ratio
  group_by(Language, position) %>%
  mutate(count = n()) %>%
  group_by(Language) %>%
  mutate(total = n()) %>%
  ungroup() %>%
  mutate(ratio = count/total) %>%
  mutate(seq = w)

data.sample <- rbind(data.sample, tmp)

}
glimpse(data.sample)
```

for each sentence extract the order of SVO
```{r}
# if needed, read the output from the previous chunks
data <- read.csv("data_tidy/UD.csv")

# get the ratio
data <- data %>%
  # only keep the words that are subjects and objects
  filter(Role %in% c("nsubj","obj")) %>%
  # take relevant columns
  select(ID_sentence, ID_word, Dependency, Role, Language) %>%
  # change columns to numeric if needed
  mutate(Dependency = as.numeric(Dependency),
         ID_word = as.numeric(ID_word)) %>%
  # get the relative position
  mutate(position = ID_word-Dependency) %>% 
  mutate(position = case_when(position < 0 ~ "before_verb",
                              position > 0 ~ "after_verb")) %>%
  # add the ratio
  group_by(Language, position) %>%
  mutate(count = n()) %>%
  group_by(Language) %>%
  mutate(total = n()) %>%
  ungroup() %>%
  mutate(ratio = count/total)
```



Plot the results with all the data
```{r}
# can change between data or data sample
data %>%
  # take the ratio of S/O before verb
  filter(position == "before_verb") %>%
  # take relevant columns
  select(Language, Role, ratio) %>%
  # remove duplicates
  distinct() %>%
  # change the format to wide
  pivot_wider(names_from = Role, values_from = ratio) %>%
  # annotate values from WALS
  mutate(Order = case_when(Language == "Arabic" ~ "VSO",
                           Language == "Basque" ~ "SOV",
                           Language == "Bulgarian" ~ "SVO",
                           Language == "Catalan" ~ "SVO",
                           Language == "Chinese" ~ "SVO",
                           Language == "Croatian" ~ "SVO",
                           Language == "Danish" ~ "SVO",
                           Language == "Dutch" ~ "No dominant",
                           Language == "English" ~ "SVO",
                           Language == "Estonian" ~ "SVO",
                           Language == "Finnish" ~ "SVO",
                           Language == "French" ~ "SVO",
                           Language == "Galician" ~ "",
                           Language == "German" ~ "No dominant",
                           Language == "Hebrew" ~ "SVO",
                           Language == "Hindi" ~ "SOV",
                           Language == "Hungarian" ~ "No dominant",
                           Language == "Indonesian" ~ "SVO",
                           Language == "Italian" ~ "SVO",
                           Language == "Japanese" ~ "SOV",
                           Language == "Korean" ~ "SOV",
                           Language == "Latin" ~ "",
                           Language == "Latvian" ~ "SVO",
                           Language == "Norwegian" ~ "SVO",
                           Language == "Persian" ~ "SOV",
                           Language == "Polish" ~ "SVO",
                           Language == "Portuguese" ~ "SVO",
                           Language == "Romanian" ~ "SVO",
                           Language == "Russian" ~ "SVO",
                           Language == "Serbian" ~ "SVO",
                           Language == "Slovak" ~ "",
                           Language == "Slovenian" ~ "SVO",
                           Language == "Spanish" ~ "SVO",
                           Language == "Swedish" ~ "SVO",
                           Language == "Ukrainian" ~ "SVO"
                           )) %>%
  # remove not used languages
  filter(!Language %in% c("Galician", "Latin", "Slovak")) %>%
  # rename columns for plot
  rename(subject_before_verb = nsubj, object_before_verb = obj) %>%
  # make the plot
  ggplot(aes(x = subject_before_verb, y= object_before_verb)) +
  geom_text(aes(label = Language, color = Order), size = 4) +
  # theme settings
  theme_bw() +
  theme(legend.position = "top",
        legend.text = element_text(size = 12)) +
  # set length of x and y axes
  xlim(c(0,1)) +
  ylim(c(0,1))
#ggsave("order_200.png", dpi = 300, width = 8, height = 6)
```




